---
title: "Tissue swab result analysis FINAL"
author: "Luka Mocivnik"
date: "2023-11-14"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE)
```



Comparison of tissue and swab samples from Proteus.

# Library import.

```{r}
library(Rsamtools)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
```

# File Import

Paths to bam files (bowtie2) and blast result tables.

```{r}
bamPath = "./data/tissue_swabs_comparisons/bowtie"
blastPath = "./data/tissue_swabs_comparisons/blast"
```


Getting names for bam and index files and importing bam files along with their indexes into a list.

```{r}
bamFiles = list.files(bamPath, pattern = ".*bam", full.names = TRUE)
bamIndex = list.files(bamPath, pattern = ".*bai", full.names = TRUE)
bamIndex = gsub(pattern = "()\\.bai", replacement = "\\1", x = bamIndex)


bamNames = list.files(bamPath, pattern = ".*bam")
bamNames = gsub(pattern = "()\\.bam", replacement = "\\1", x = bamNames)

bamImport = lapply(1:length(bamFiles), function(x, b, i){
  out = scanBam(b[x], index = i[x])
},b = bamFiles, i = bamIndex)

names(bamImport) = bamNames

```


Extracting query and subject sequence names from bam list.

```{r}

querySubjectBowtie = lapply(1:length(bamImport), function(x, bam_list, comboNames){
  qname = bam_list[[x]][[1]]$qname
  rname = as.character(bam_list[[x]][[1]]$rname)
  
  df = data.frame(qname, rname)

  df$qname = as.numeric(df$qname)
  
  #for mapping against the genome we add start position to scaffold name in order to increase the number of subject_ids, otherwise results become weird and incomparable to non genome mapping results
  isNumber = sum(!is.na(as.numeric(df$rname)))
  
  if (isNumber > 0) {
    df$rname = as.numeric(df$rname)
  }else{
    #adding start number in position to increase unique hits
    pos = as.character(bam_list[[x]][[1]]$pos)
    df$rname = paste(df$rname, pos, sep = "_")
  }
  df$rname = na_if(as.character(df$rname), "NA_NA")
  df$unique_subject = !df$rname %in% df$rname[duplicated(df$rname)]

  
  colnames(df) = c("query_id", paste("subject", comboNames[[x]], sep = "_"), paste("unique_subject", comboNames[[x]], sep = "_"))
  df
  
}, bam_list = bamImport, comboNames = bamNames)

names(querySubjectBowtie) = bamNames

```

Reading blast results and taking query and subject columns. This should be fairly straightforward.

```{r}
blastFiles = list.files(blastPath, pattern = "blast.*csv", full.names = TRUE)
blastNames = list.files(blastPath, pattern = "blast.*csv")
blastNames = gsub(pattern = "()_Top.*", replacement = "\\1", x = blastNames)

blastTopResults = lapply(blastFiles, read.csv, header = TRUE)
names(blastTopResults) = blastNames

querySubjectBlast = lapply(blastTopResults, function(x){
  
  #combine scaffold number and hit start in genome comparisons, similar to the bowtie2 data
  if (class(x$subject_id) == "character") {
    x$subject_id = paste(x$subject_id, as.character(x$s_start), sep ="_")
  }
  
  
  out = x[,1:2]
  out$unique_subject_blast = !out$subject_id %in% out$subject_id[duplicated(out$subject_id)]
  out
})

```

# Joining results

Joining results into tables by different combinations: whole tissue vs. swabs, whole swabs vs tissue, split swabs and split tissue vs. whole swabs and tissue and split swabs and tissue vs split swabs and tissue. Six combinations in total. Genome data was added later in the same manner, four extra tables, 10 in total. Whole tables are sent separately, they are a little too large for the report.

```{r}
querySubjectAllSvT = querySubjectBowtie$bowtie2AllSvT_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2AllSvT_local_sorted) %>%
  left_join(querySubjectBlast$blastAllSvT)

querySubjectAllSvT$unique_subject_bowtie2AllSvT_global_sorted[is.na(querySubjectAllSvT$subject_bowtie2AllSvT_global_sorted)] = NA
querySubjectAllSvT$unique_subject_bowtie2AllSvT_local_sorted[is.na(querySubjectAllSvT$subject_bowtie2AllSvT_local_sorted)] = NA

querySubjectAllTvS = querySubjectBowtie$bowtie2AllTvS_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2AllTvS_local_sorted) %>%
  left_join(querySubjectBlast$blastAllTvS)

querySubjectAllTvS$unique_subject_bowtie2AllTvS_global_sorted[is.na(querySubjectAllTvS$subject_bowtie2AllTvS_global_sorted)] = NA
querySubjectAllTvS$unique_subject_bowtie2AllTvS_local_sorted[is.na(querySubjectAllTvS$subject_bowtie2AllTvS_local_sorted)] = NA

querySubjectSplitAllSvT = querySubjectBowtie$bowtie2SplitAllSvT_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2SplitAllSvT_local_sorted) %>%
  left_join(querySubjectBlast$blastSplitAllSvT)

querySubjectSplitAllSvT$unique_subject_bowtie2SplitAllSvT_global_sorted[is.na(querySubjectSplitAllSvT$subject_bowtie2SplitAllSvT_global_sorted)] = NA
querySubjectSplitAllSvT$unique_subject_bowtie2SplitAllSvT_local_sorted[is.na(querySubjectSplitAllSvT$subject_bowtie2SplitAllSvT_local_sorted)] = NA

querySubjectSplitAllTvS = querySubjectBowtie$bowtie2SplitAllTvS_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2SplitAllTvS_local_sorted) %>%
  left_join(querySubjectBlast$blastSplitAllTvS)

querySubjectSplitAllTvS$unique_subject_bowtie2SplitAllTvS_global_sorted[is.na(querySubjectSplitAllTvS$subject_bowtie2SplitAllTvS_global_sorted)] = NA
querySubjectSplitAllTvS$unique_subject_bowtie2SplitAllTvS_local_sorted[is.na(querySubjectSplitAllTvS$subject_bowtie2SplitAllTvS_local_sorted)] = NA

querySubjectSplitSvT = querySubjectBowtie$bowtie2SplitSvT_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2SplitSvT_local_sorted) %>%
  left_join(querySubjectBlast$blastSplitSvT)

querySubjectSplitSvT$unique_subject_bowtie2SplitSvT_global_sorted[is.na(querySubjectSplitSvT$subject_bowtie2SplitSvT_global_sorted)] = NA
querySubjectSplitSvT$unique_subject_bowtie2SplitSvT_local_sorted[is.na(querySubjectSplitSvT$subject_bowtie2SplitSvT_local_sorted)] = NA

querySubjectSplitTvS = querySubjectBowtie$bowtie2SplitTvS_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2SplitTvS_local_sorted) %>%
  left_join(querySubjectBlast$blastSplitTvS)

querySubjectSplitTvS$unique_subject_bowtie2SplitTvS_global_sorted[is.na(querySubjectSplitTvS$subject_bowtie2SplitTvS_global_sorted)] = NA
querySubjectSplitTvS$unique_subject_bowtie2SplitTvS_local_sorted[is.na(querySubjectSplitTvS$subject_bowtie2SplitTvS_local_sorted)] = NA

#adding genome data

querySubjectAllSvG = querySubjectBowtie$bowtie2AllSvG_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2AllSvG_local_sorted) %>%
  left_join(querySubjectBlast$blastAllSvG)

querySubjectAllSvG$unique_subject_bowtie2AllSvG_global_sorted[is.na(querySubjectAllSvG$subject_bowtie2AllSvG_global_sorted)] = NA
querySubjectAllSvG$unique_subject_bowtie2AllSvG_local_sorted[is.na(querySubjectAllSvG$subject_bowtie2AllSvG_local_sorted)] = NA

querySubjectAllTvG = querySubjectBowtie$bowtie2AllTvG_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2AllTvG_local_sorted) %>%
  left_join(querySubjectBlast$blastAllTvG)

querySubjectAllTvG$unique_subject_bowtie2AllTvG_global_sorted[is.na(querySubjectAllTvG$subject_bowtie2AllTvG_global_sorted)] = NA
querySubjectAllTvG$unique_subject_bowtie2AllTvG_local_sorted[is.na(querySubjectAllTvG$subject_bowtie2AllTvG_local_sorted)] = NA

querySubjectSplitSvG = querySubjectBowtie$bowtie2SplitSvG_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2SplitSvG_local_sorted) %>%
  left_join(querySubjectBlast$blastSplitSvG)

querySubjectSplitSvG$unique_subject_bowtie2SplitSvG_global_sorted[is.na(querySubjectSplitSvG$subject_bowtie2SplitSvG_global_sorted)] = NA
querySubjectSplitSvG$unique_subject_bowtie2SplitSvG_local_sorted[is.na(querySubjectSplitSvG$subject_bowtie2SplitSvG_local_sorted)] = NA

querySubjectSplitTvG = querySubjectBowtie$bowtie2SplitTvG_global_sorted %>%
  left_join(querySubjectBowtie$bowtie2SplitTvG_local_sorted) %>%
  left_join(querySubjectBlast$blastSplitTvG)

querySubjectSplitTvG$unique_subject_bowtie2SplitTvG_global_sorted[is.na(querySubjectSplitTvG$subject_bowtie2SplitTvG_global_sorted)] = NA
querySubjectSplitTvG$unique_subject_bowtie2SplitTvG_local_sorted[is.na(querySubjectSplitTvG$subject_bowtie2SplitTvG_local_sorted)] = NA

```


# Basic metrics and graphs

Some basic metrics on joined data.

```{r}
#making a list of all the combinations
querySubjectList = list(querySubjectAllSvT, querySubjectAllTvS, querySubjectSplitAllSvT, querySubjectSplitAllTvS, querySubjectSplitSvT, querySubjectSplitTvS, querySubjectAllSvG, querySubjectAllTvG, querySubjectSplitSvG, querySubjectSplitTvG)
names(querySubjectList) = c("querySubjectAllSvT", "querySubjectAllTvS", "querySubjectSplitAllSvT", "querySubjectSplitAllTvS", "querySubjectSplitSvT", "querySubjectSplitTvS", "querySubjectAllSvG", "querySubjectAllTvG", "querySubjectSplitSvG", "querySubjectSplitTvG")

#checking where different mapping settings and software produced the same results
querySubjectList = lapply(querySubjectList, function(x){
  x$bt2_global_local_same = x[,2] == x[,4] | (is.na(x[,2]) & is.na(x[,4]))
  x$bt2_global_local_same[is.na(x$bt2_global_local_same)] = FALSE

  x$bt2_global_blast_same = x[,2] == x[,6] | (is.na(x[,2]) & is.na(x[,6]))
  x$bt2_global_blast_same[is.na(x$bt2_global_blast_same)] = FALSE

  x$bt2_local_blast_same = x[,4] == x[,6] | (is.na(x[,4]) & is.na(x[,6]))
  x$bt2_local_blast_same[is.na(x$bt2_local_blast_same)] = FALSE
  
  x$all_same = x[,2] == x[,4] & x[,2] == x[,6] & x[,4] == x[,6] | (is.na(x[,2]) & is.na(x[,4]) & is.na(x[,6]))
  x$all_same[is.na(x$all_same)] = FALSE
  
  x
  
})


querySubjectList_basicMetrics = lapply(querySubjectList, function(x){
  
  #percentage of unique hits with and without NAs (lines with NAs are ore are not taken into account when calculating the percentage)
  out = data.frame(Bt2_perc_global_uniq_withNA = sum(x[,3], na.rm = TRUE)/nrow(x))
  out$Bt2_perc_local_uniq_withNA = sum(x[,5], na.rm = TRUE)/nrow(x)
  out$blast_perc_uniq_withNA = sum(x[,7], na.rm = TRUE)/nrow(x)
  out$Bt2_perc_global_uniq_woNA = sum(x[,3], na.rm = TRUE)/(nrow(x)-sum(is.na(x[,3])))
  out$Bt2_perc_local_uniq_woNA = sum(x[,5], na.rm = TRUE)/(nrow(x)-sum(is.na(x[,5])))
  out$blast_perc_uniq_woNA = sum(x[,7], na.rm = TRUE)/(nrow(x)-sum(is.na(x[,7])))

  
  #percentage of all, not just unique hits, NAs included
  out$Bt2_perc_hits_global = 1-(sum(is.na(x[,2]))/nrow(x))
  out$Bt2_perc_hits_global = 1-(sum(is.na(x[,2]))/nrow(x))
  
  out$Bt2_perc_hits_local = 1-(sum(is.na(x[,4]))/nrow(x))
  out$blast_perc_hits = 1-(sum(is.na(x[,6]))/nrow(x))


  #percentage of sequences hitting the same subject_id with different mapping methods
  out$perc_Bt2_global_local_same = sum(x[,8])/nrow(x)
  out$perc_Bt2_global_blast_same = sum(x[,9])/nrow(x)
  out$perc_Bt2_local_blast_same = sum(x[,10])/nrow(x)
  out$perc_all_same = sum(x[,11])/nrow(x)
  
  #added later
  out$Bt2_perc_local_NOTuniq_withNA = sum(x[,5] == FALSE, na.rm = TRUE)/nrow(x)
  out$Bt2_perc_local_NA_withNA = sum(is.na(x[,5]))/nrow(x)
  
  #absolute numbers
  out$Bt2_perc_global_uniq_NUM = sum(x[,3], na.rm = TRUE)
  out$Bt2_perc_global_NUM = nrow(x) - sum(is.na(x[,3]))

  out$Bt2_perc_local_uniq_NUM = sum(x[,5], na.rm = TRUE)
  out$Bt2_perc_local_NUM = nrow(x) - sum(is.na(x[,5]))
  
  out$blast_perc_uniq_NUM = sum(x[,7], na.rm = TRUE)
  out$blast_perc_NUM = nrow(x) - sum(is.na(x[,7]))
  
  out
})

#used to get different metrics
querySubjectMetrics = do.call(rbind, querySubjectList_basicMetrics)
querySubjectMetricsWITH_NUMS = querySubjectMetrics

querySubjectMetrics = querySubjectMetrics[,c(1:15)]
rownames(querySubjectMetrics) = gsub("querySubject(.*)", replacement = "\\1", x = rownames(querySubjectMetrics))
querySubjectMetrics$names = rownames(querySubjectMetrics)

# making the final table
querySubjectMetricsFINAL = querySubjectMetrics %>%
  filter(names %in% c("AllSvT", "SplitAllSvT", "SplitSvT", "AllSvG")) %>%
  mutate(names = gsub(pattern = "^AllSvT$", replacement = "swab vs. tissue full", x = names)) %>%
  mutate(names = gsub(pattern = "^SplitAllSvT$", replacement = "swab split vs. tissue", x = names)) %>%
  mutate(names = gsub(pattern = "^SplitSvT$", replacement = "swab split vs. tissue split", x = names)) %>%
  mutate(names = gsub(pattern = "^AllSvG$", replacement = "swabs vs. genome", x = names)) %>%
  t()

querySubjectMetricsFINAL = as.data.frame(querySubjectMetricsFINAL)
colnames(querySubjectMetricsFINAL) = querySubjectMetricsFINAL["names",]
querySubjectMetricsFINAL = querySubjectMetricsFINAL[1:(nrow(querySubjectMetricsFINAL)-3),]

querySubjectMetricsFINAL = apply(querySubjectMetricsFINAL, 2, as.numeric)

querySubjectMetricsFINAL = apply(querySubjectMetricsFINAL, 2, round, 3)
rownames(querySubjectMetricsFINAL) = c("unique with NAs bowtie gl.", "unique with NAs bowtie lo.", "unique with NAs blast", "unique without NAs bowtie gl.", "unique without NAs bowtie lo.", "unique without NAs blast", "total bowtie gl.", "total bowtie lo.", "total blast", "overlapping bowtie gl. vs lo.", "overlapping bowtie gl. vs blast", "overlapping lo. vs blast", "all similar")


```

Writing results to disk.

```{r, eval = F}
write.csv(querySubjectMetricsWITH_NUMS, file = "proteus_swab_tissue_metricsWITH_NUMS.csv")
sapply(names(querySubjectList), function(x){write.csv(querySubjectList[[x]], file=paste(x, "csv", sep="."))})
```


Drawing some results.

```{r}


#percentage of total hits

table_perc_hits = pivot_longer(querySubjectMetrics[, c(1:13, 16)], cols = c(7:9), names_to = "metric", values_to = "values") %>%
  select(11:13)

#name replacements
table_perc_hitsFINAL = table_perc_hits %>%
  filter(names %in% c("AllSvT", "SplitSvT", "SplitAllSvT")) %>%
  mutate(metric = (gsub(pattern = "^Bt2_perc_hits_global$", replacement = "Bowtie2 global", x = metric))) %>%
  mutate(metric = (gsub(pattern = "^Bt2_perc_hits_local$", replacement = "Bowtie2 local", x = metric))) %>%
  mutate(metric = (gsub(pattern = "^blast_perc_hits$", replacement = "Blast", x = metric)))

# ggplot(table_perc_hits, aes(x = names, y = values, fill = metric)) +
#   geom_bar(position = "dodge", stat = "identity") +
#   scale_x_discrete(guide = guide_axis(angle = 45)) +
#   #scale_y_continuous(n.breaks=6) +
#   ylim(0:1) +
#   labs(title = "Percentage of found sequences", x = "Sequence concatenation", y = "Proportion of mapped loci")

ggplot(table_perc_hitsFINAL, aes(x = names, y = values, fill = metric)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  #scale_y_continuous(n.breaks=6) +
  ylim(0:1) +
  labs(title = "Percentage of found sequences", x = "Sequence concatenation", y = "Proportion of mapped loci")


```

Plots commissioned by Hans.

```{r}

#best method SvT and SvG


querySubjectMetrics_bestMethodTG = querySubjectMetrics %>%
  dplyr::filter(names == "AllSvT" | names == "AllSvG") %>%
  pivot_longer(cols = c(1:9, 14,15), names_to = "metric", values_to = "values") %>%
  select(c("names", "metric", "values")) %>%
  dplyr::filter(grepl(pattern = "local", x = metric)) %>%
  mutate(prettyMetric = metric) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_uniq_withNA", replacement = "unique matches", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_uniq_woNA", replacement = "Unique percentage without unmtched", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_hits_local", replacement = "Percentage All", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_NOTuniq_withNA", replacement = "non-unique matches", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_NA_withNA", replacement = "unmatched", x = prettyMetric)) %>%
  group_by(names) %>%
  arrange(values, .by_group = TRUE) %>%
  mutate(prettyMetric = fct_inorder(factor(prettyMetric)))


# querySubjectMetrics_bestMethodTG %>%
#   filter(grepl(".*withNA$", metric)) %>%
#   ggplot(aes(x = names, y = values, fill = prettyMetric)) +
#     geom_bar(position = "dodge", stat = "identity") +
#     scale_x_discrete(guide = guide_axis(angle = 45)) +
#     #scale_y_continuous(n.breaks=6) +
#     ylim(0:1) +
#     labs(y = "Proportion")

#ggsave(filename = paste0("SvT", "_", "found_sequences_typesTG", ".eps"), path = "./figures")



#adding TvG
querySubjectMetrics_bestMethodSTG = querySubjectMetrics %>%
  dplyr::filter( names == "AllSvG" | names == "AllTvG") %>%
  pivot_longer(cols = c(1:9, 14,15), names_to = "metric", values_to = "values") %>%
  select(c("names", "metric", "values")) %>%
  dplyr::filter(grepl(pattern = "local", x = metric)) %>%
  mutate(prettyMetric = metric) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_uniq_withNA", replacement = "unique matches", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_uniq_woNA", replacement = "Unique percentage without unmtched", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_hits_local", replacement = "Percentage All", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_NOTuniq_withNA", replacement = "non-unique matches", x = prettyMetric)) %>%
  mutate(prettyMetric = gsub(pattern = "Bt2_perc_local_NA_withNA", replacement = "unmatched", x = prettyMetric)) %>%
  group_by(names) %>%
  arrange(values, .by_group = TRUE) %>%
  mutate(prettyMetric = fct_inorder(factor(prettyMetric)))

# querySubjectMetrics_bestMethodSTG %>%
#   filter(grepl(".*withNA$", metric)) %>%
#   ggplot(aes(x = names, y = values, fill = prettyMetric)) +
#     geom_bar(position = "dodge", stat = "identity") +
#     scale_x_discrete(guide = guide_axis(angle = 45)) +
#     #scale_y_continuous(n.breaks=6) +
#     ylim(0:1) +
#     labs(y = "Proportion")

#ggsave(filename = paste0("SvG_TvG", "_", "genome_comparison", ".eps"), path = "./figures")


#chart for paper
querySubjectMetrics_bestMethodFINAL = querySubjectMetrics_bestMethodSTG %>%
  filter(names == "AllTvG") %>%
  bind_rows(querySubjectMetrics_bestMethodTG) %>%
  mutate(names = (gsub(pattern = "^AllTvG$", replacement = "tissue catalogue vs. genome", x = names))) %>%
  mutate(names = (gsub(pattern = "^AllSvG$", replacement = "swab catalogue vs. genome", x = names))) %>%
  mutate(names = (gsub(pattern = "^AllSvT$", replacement = "swab vs. tissue catalogue", x = names)))
  
  

querySubjectMetrics_bestMethodFINAL %>%
  filter(grepl(".*withNA$", metric)) %>%
  ggplot(aes(x = names, y = values, fill = prettyMetric)) +
    geom_bar(position = "dodge", stat = "identity") +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    #scale_y_continuous(n.breaks=6) +
    ylim(0:1) +
    labs(y = "Proportion")




```


