---
title: "Tissue swabs individual taxa analysis Megan"
output: html_document
---

Library import.

```{r}
library(stringr)
library(dplyr)
library(tidyr)
```

Getting a list of Megan output files for import.

```{r}
taxaTablesFullNames = list.files(path = "./path/to/megan/output/", full.names = TRUE)
taxaTablesNames = list.files(path = "./path/to/megan/output/")

taxaTableList = lapply(taxaTablesFullNames, read.table, sep = "\t")
names(taxaTableList) = taxaTablesNames

```

Importing data

```{r}
intxtdf = lapply(taxaTablesFullNames, function(x){

  out = readLines(x)

  out = out[6:(length(out) - 1)]
  outdf = as.data.frame(out)
  outdf = as.data.frame(str_split_fixed(outdf$out, pattern = ": ", 2))
  outdf$V1 = str_squish(outdf$V1)
  colnames(outdf) = c("taxon", "count")
  return(outdf)

})
names(intxtdf) = taxaTablesNames

intxtdf2 = lapply(1:length(intxtdf), function(x, inlist){
  
  df = inlist[[x]]
  dfname = names(inlist[x])
  dfsample = gsub(pattern = "(PC\\d{3}_[ts])_.*", replacement = "\\1", x = dfname)
  df$name = dfsample
  return(df)
  
}, inlist = intxtdf)

names(intxtdf2) = taxaTablesNames

```

Getting data for taxonomic ranks of interest separated.

```{r}
intxtdfSuperkingdoms = intxtdf2[grep(pattern = "superkingdom", x = names(intxtdf2))]
intxtdfBacteriaOrder = intxtdf2[grep(pattern = "bacteriaOrder", x = names(intxtdf2))]
intxtdfVerebrateClass = intxtdf2[grep(pattern = "vertebrateClass", x = names(intxtdf2))]

```

Using a separate function to change lists in data frames with ratios.

```{r}
source("./scripts/megan_individual_taxa_df.R")
dfSuperkingdoms = megan_individual_taxa(intxtdfSuperkingdoms)
dfBacteriaOrder = megan_individual_taxa(intxtdfBacteriaOrder)
dfVertebrateClass = megan_individual_taxa(intxtdfVerebrateClass)

```
Getting the Other category for bacterial orders.

```{r}
dfBacteriaOrder_otherCol = apply(dfBacteriaOrder[,2:(ncol(dfBacteriaOrder)-1)], c(1,2), function(x){
  out = x
  if (is.na(x)) {
    out = NA
  }else if(x >= 0.05){
    out = NA
  }else{
    out = x
  }
  return(out)
})
dfBacteriaOrder_otherCol = apply(dfBacteriaOrder_otherCol, 2, as.numeric)
Other = rowSums(dfBacteriaOrder_otherCol, na.rm = TRUE)

dfBacteriaOrderOut1 = apply(dfBacteriaOrder[,2:(ncol(dfBacteriaOrder)-1)], c(1,2), function(x){
  out = x
  if (is.na(x)) {
    out = NA
  }else if(x < 0.05){
    out = NA
  }else{
    out = x
  }
  return(out)
})

dfBacteriaOrderOut = bind_cols(dfBacteriaOrder[,1], dfBacteriaOrderOut1, Other = Other, Sample_type = dfBacteriaOrder$Sample_type)
```

Writing tables. Change eval to TRUE to run this chunk.

```{r, eval = F}
write.csv(dfBacteriaOrderOut, file = "megan_bacteria_order_limit.csv")
write.csv(dfVertebrateClass, file = "megan_chordata_class_limit.csv")
write.csv(dfSuperkingdoms, file = "megan_eukaryota_superkingdoms_limit.csv")
```

Counting percentages for specific categories from the tables 

```{r}

#read count
swabAllseq = 88577446
tissueAllseq = 340773560

#bacteria
swabPercBacteria = bind_rows(intxtdfBacteriaOrder) %>%
  filter(grepl(pattern = "*_s$", x = name) == TRUE) %>%
  mutate(count = gsub(pattern = ",", replacement = "", x = count)) %>%
  mutate(count = as.integer(count)) %>%
  summarise(percBac = (sum(count)/swabAllseq)*100) %>%
  as.numeric(.)

tissuePercBacteria = bind_rows(intxtdfBacteriaOrder) %>%
  filter(grepl(pattern = "*_t$", x = name) == TRUE) %>%
  mutate(count = gsub(pattern = ",", replacement = "", x = count)) %>%
  mutate(count = as.integer(count)) %>%
  summarise(percBac = (sum(count)/tissueAllseq)*100) %>%
  as.numeric(.)

#fish
swabPercFish = bind_rows(intxtdfVerebrateClass) %>%
  filter(grepl(pattern = "*_s$", x = name) == TRUE) %>%
  filter(taxon == "Actinopteri") %>%
  mutate(count = gsub(pattern = ",", replacement = "", x = count)) %>%
  mutate(count = as.integer(count)) %>%
  summarise(percBac = (sum(count)/swabAllseq)*100) %>%
  as.numeric(.)

tissuePercFish = bind_rows(intxtdfVerebrateClass) %>%
  filter(grepl(pattern = "*_t$", x = name) == TRUE) %>%
  filter(taxon == "Actinopteri") %>%
  mutate(count = gsub(pattern = ",", replacement = "", x = count)) %>%
  mutate(count = as.integer(count)) %>%
  summarise(percBac = (sum(count)/tissueAllseq)*100) %>%
  as.numeric(.)

#from megan
swabHumanHits = 1193
tissueHUmanHits = 3831

swabHumanPerc = (swabHumanHits / swabAllseq)*100
tissueHUmanPerc = (tissueHUmanHits / tissueAllseq)*100
```



